---
title: "Creating raster mask from EEZs of the world shapefile"
author: "Denisse Fierro Arcos"
date: "2022-11-07"
output: 
  github_document:
    toc: true
    html_preview: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

The EEZs the of the world shapefile came from [VLIZ](https://doi.org/10.14284/386). Rasters are created in `R` and the final details are done with `xarray` library for `Python`.

## Loading R libraries

```{r libraries, results = "hide", warnings = F, message = F}
library(sf)
library(raster)
library(tidyverse)
library(reticulate)
```

## Loading shapefiles
The original shapefile contained a couple of errors:
1. The country of Mauritius appeared as both Mauritius and Republic of Mauritius, however they both referred to the same country as their country code was the same for both: MUS.
2. The United Kingdom appeared twice with two different countries codes. GBR has been kept as the unique code to represent this country.
  
The section below loads the file, selects a subset of columns of interest and corrects the errors identified above.

```{r shapefiles_dirs}
#Load shapefile with EEZ
eez_world <- read_sf("../Data/World_EEZ_v11_20191118/eez_v11.shp") %>%
   #Subset of columns
  select(MRGID, GEONAME, POL_TYPE, SOVEREIGN1, SOVEREIGN2, SOVEREIGN3, AREA_KM2) %>% 
  #Turning character columns into factors
  mutate_if(is.character, as.factor)

#We can check the results of the first two rows
head(eez_world, 2)

#Identify EEZ and countries associated with them from extracted data files
eez_names_codes <- eez_world %>% 
  st_drop_geometry()
```

Saving the shapefile and keys for reference.
```{r eval = F}
#Saving the corrected shapefile
eez_world %>% 
  st_write("../Data/World_EEZ_v11_20191118/eez_short.shp", delete_layer = T)

#Saving keys
eez_names_codes %>% 
  write_csv("../Data/World_EEZ_v11_20191118/eez_key.csv")
```

## Plotting FAO regions shapefile
```{r plot_shapefile}
#Loading land shapefile to include in plot
land <- rnaturalearth::ne_countries(type = "countries", returnclass = "sf")

#Plotting FAO shapefile
eez_world %>% 
  ggplot()+
  geom_sf(aes(fill = SOVEREIGN1))+
  geom_sf(data = land, inherit.aes = F, color = "gray")+
  theme_bw()+
  theme(legend.position = "none")
```

## Extracting names and codes for EEZ
We will use this information to save correct names for regions in the raster

```{r eez_keys}
#Create a data frame of unique FAO regions
eez_names_codes <-  eez_names_codes %>% 
  group_by(MRGID) %>% 
  group_keys()

#We can check some of the results
head(eez_names_codes, 2)
```


## Creating a multilayer raster mask based on merged shapefile
We will now create multilayer mask, which match the resolution of the model forcings (1 deg). First, we will load the sample rasters.

### Loading input rasters
```{r load_rasters_input}
#Loading 1 deg raster to be used as target for rasterising FAO regions
deg1 <- raster("../Data/InputRasters/gfdl-mom6-cobalt2_obsclim_deptho_onedeg_global_fixed.nc")
#Sample from DBPM model
#deg1 <- raster("../Data/InputRasters/dbpm_ipsl-cm6a-lr_nobasd_historical_nat_default_tcb_global_monthly_1850_2014.nc")[[1]]
#Sample from DBEM model
#deg05 <- raster("../Data/InputRasters/dbem_ipsl-cm6a-lr_nobasd_historical_nat_default_tcb_global_annual_1951_2014.nc")[[1]]

#Plotting raster
plot(deg1)
#plot(deg05)
```
### Calculate grid area
The `raster` package allows us to calculate the area of grid cells in $km^2$ in just one line of code. We can save this to calculate weighted means by area. We will use the raster above as a base, and we will save the result on our disk.

```{r eval = F}
#Calculating grid area
deg_area <- area(deg1)

#Saving results
writeRaster(deg_area, "../Data/InputRasters/area_1deg.nc", format = "CDF", overwrite = T, 
            #We will save the variable names so they match the Fish-MIP models
            varname = "area", xname = "lon", yname = "lat", varunit = "km2")
```



### Defining function to create rasters from shapefiles
We will define our own function that will use the shapefiles above to create rasters.

```{r function_shp_raster}
#Defining function which needs a shapefile and a raster as input
shp_to_raster <- function(shp, nc_raster){
  #The final raster will have ones where within the shapefile boundaries
  rasterize(shp, nc_raster, field = 1)
}

```

### Applying function to list containing all shapefiles

```{r mapping_function}
#Split shapefile into regions prior to transforming into raster
eez_list <- eez_world %>% 
  group_by(MRGID) %>% 
  group_split()

#Applying function to raster list
deg1_raster <- map(eez_list, shp_to_raster, deg1) %>% 
  #Stacking rasters to create a single multilayer raster
  stack()

#Checking results of stacked raster (selected Australian EEZs)...
plot(deg1_raster[[c(31:34, 45, 103)]])
```

```{r, eval = F}
#Saving raster to disk
writeRaster(deg1_raster, "../Data/Masks/EEZ-world_1degmask.nc", format = "CDF", overwrite = T,
            varname = "EEZ_regions", zname = "Country_EEZ")
```

# `Python`-based code
We will now start `Python` and save the correct names for the FAO regions in the `netcdf` file we created in `R`.

```{r}
#Activating conda
use_condaenv("CMIP6_data")
```

## Loading libraries

```{python}
import xarray as xr
import numpy as np
import matplotlib.pyplot as plt
import pandas as pd
```

## Loading raster using `xarray`
```{python}
#Loading multilayer raster as dataset
mask = xr.open_dataset("../Data/Masks/EEZ-world_1degmask.nc")
#Checking saved file in R
mask
```
Here we can see that the FAO region names are not saved correctly. They are numbered based on its location on the shapefile. We can update this using the data frame with the unique FAO regions we previously created.

```{python}
#We can simply load the data frame from the R environment.
eez_keys = r.eez_names_codes

#Checking list
eez_keys
```
We can now update the names on the `netcdf` file using the ID codes for each EEZ.

```{python}
#We will use the values in the data frame we loaded above
mask['Country_EEZ'] = [int(i) for i in eez_keys.MRGID]
```

We will load the full keys into memory and we will select all EEZ areas for Australia to ensure masks are correct.

```{python}
#Loading keys
eez_full_keys = pd.read_csv('../Data/World_EEZ_v11_20191118/eez_key.csv')

#Selecting Australian subset
aus_eez = mask.sel(Country_EEZ = eez_full_keys['MRGID'][eez_full_keys['SOVEREIGN1'] == 'Australia'].tolist()).EEZ_regions
```


## Plotting results
We will plot all regions below to ensure we got them all correctly.
```{python}
#We will loop through the first five to check results
for reg in aus_eez:
  #Plotting results
  fig = plt.figure()
  ax = fig.add_subplot(111)
  reg.plot(ax = ax, levels = [1, 2])
  plt.title(reg.Country_EEZ.values.tolist())
  plt.show()

```
When comparing to the shapefile plots in the `R` section of this notebook, we can see that the regions are named correctly. This means that we can save our results now.

```{python, eval = F}
mask.to_netcdf("../Data/Masks/EEZ-world-corrected_1degmask.nc")
```

