---
title: "Creating raster mask from EEZs of the world shapefile"
author: "Denisse Fierro Arcos"
date: "2022-11-07"
output: 
  github_document:
    toc: true
    html_preview: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

The EEZs the of the world shapefile came from [VLIZ](https://doi.org/10.14284/386). Rasters are created in `R` and the final details are done with `xarray` library for `Python`.

## Loading R libraries

```{r libraries, results = "hide", warnings = F, message = F}
library(sf)
library(raster)
library(tidyverse)
library(reticulate)
```

## Loading shapefiles
The original shapefile contained a couple of errors:
1. The country of Mauritius appeared as both Mauritius and Republic of Mauritius, however they both referred to the same country as their country code was the same for both: MUS.
2. The United Kingdom appeared twice with two different countries codes. GBR has been kept as the unique code to represent this country.
  
The section below loads the file, selects a subset of columns of interest and corrects the errors identified above.

```{r shapefiles_dirs}
#Load shapefile with EEZ
eez_world <- read_sf("../Spatial_Data/World_EEZ_v11_20191118/eez_v11.shp") %>%
   #Subset of columns
  select(MRGID, GEONAME, POL_TYPE, SOVEREIGN1, SOVEREIGN2, SOVEREIGN3, AREA_KM2) %>% 
  #Turning character columns into factors
  mutate_if(is.character, as.factor)

#We can check the results of the first two rows
head(eez_world, 2)

#Identify EEZ and countries associated with them from extracted data files
eez_names_codes <- eez_world %>% 
  st_drop_geometry()
```

Saving the shapefile and keys for reference.
```{r eval = F}
#Saving the corrected shapefile
eez_world %>% 
  st_write("../Spatial_Data/World_EEZ_v11_20191118/eez_short.shp", delete_layer = T)

#Saving keys
eez_names_codes %>% 
  write_csv("../Spatial_Data/World_EEZ_v11_20191118/eez_key.csv")
```

## Plotting FAO regions shapefile
```{r plot_shapefile}
#Loading land shapefile to include in plot
land <- rnaturalearth::ne_countries(type = "countries", returnclass = "sf")

#Plotting FAO shapefile
eez_world %>% 
  ggplot()+
  geom_sf(aes(fill = SOVEREIGN1))+
  geom_sf(data = land, inherit.aes = F, color = "gray")+
  theme_bw()+
  theme(legend.position = "none")
```

## Extracting names and codes for EEZ
We will use this information to save correct names for regions in the raster

```{r eez_keys}
#Create a data frame of unique FAO regions
eez_names_codes <-  eez_names_codes %>% 
  group_by(MRGID) %>% 
  group_keys()

#We can check some of the results
head(eez_names_codes, 2)
```


## Creating a multilayer raster mask based on merged shapefile
We will now create a multilayer mask that matches the grid used in the physical model forcings. Most models use the same 1 degree grid, with the exception of the DBPM ecosystem model, which uses a different 1 degree grid, and the DBEM model, which uses a 0.5 degree grid.  
  
Note that masks *must* match the grid of the model from which data is being extracted. This means that you will need to create a new mask for each grid that is different. In the chunk below, you will find the three different grids identified in the ecosystem models.

### Loading input rasters

```{r load_rasters_input}
#Loading sample raster to be used as target for rasterising FAO regions
#Most ecosystem models use this one degree grid
ras <- raster("../Spatial_Data/InputRasters/gfdl-mom6-cobalt2_obsclim_deptho_1deg_global_fixed.nc")

#Sample from DBPM model
# ras <- raster("../Spatial_Data/InputRasters/dbpm_ipsl-cm6a-lr_nobasd_historical_nat_default_1deg_tcb_global_monthly_1850_2014.nc")[[1]]

#Sample from DBEM model
# ras <- raster("../Spatial_Data/InputRasters/dbem_ipsl-cm6a-lr_nobasd_historical_nat_default_05deg_tcb_global_annual_1951_2014.nc")[[1]]

#Sample for 0.25 degree grid
# ras <- raster("../Spatial_Data/InputRasters/gfdl-mom6-cobalt2_obsclim_deptho_025deg_global_fixed.nc")

#We will define a few extra variables to automate creation of file names for each mask
#Model resolution
res <- "1deg"

#Model associated with grid. Leave blank if multiple models use same grid
mod_name <- ""

#Plotting raster
plot(ras)
```
### Calculate grid area
The `raster` package allows us to calculate the area of grid cells in $km^2$ in just one line of code. We can save this to calculate weighted means by area. We will use the raster above as a base, and we will save the result on our disk.

```{r eval = F}
#Calculating grid area
area_grid <- area(ras)

#Creating file name 
filename <- paste0("area_", res, mod_name, ".nc")

#Saving results
writeRaster(area_grid, file.path("../Spatial_Data/Masks", filename), format = "CDF", overwrite = T, 
            #We will save the variable names so they match the Fish-MIP models
            varname = "area", xname = "lon", yname = "lat", varunit = "km2")
```

### Defining function to create rasters from shapefiles
We will define our own function that will use the shapefiles above to create rasters.

```{r function_shp_raster}
#Defining function which needs a shapefile and a raster as input
shp_to_raster <- function(shp, nc_raster){
  #The final raster will have ones where within the shapefile boundaries
  rasterize(shp, nc_raster, field = 1)
}

```

### Applying raster creation function
We will apply the function we defined above to all the polygons contained within our shapefile. We can do this easily using the `map` function from the `purrr` package.

```{r mapping_function}
#Split shapefile into regions prior to transforming into raster
eez_list <- eez_world %>% 
  group_by(MRGID) %>% 
  group_split()

#Applying function to raster list
grid_raster <- map(eez_list, shp_to_raster, ras) %>% 
  #Stacking rasters to create a single multilayer raster
  stack()

#Checking results of stacked raster (selected Australian EEZs)...
plot(grid_raster[[c(31:34, 45, 103)]])
```

Finally, we save the outputs in our local machine.
```{r, eval = F}
#Define file name
filename <- paste0("EEZ-world_", res, "mask", mod_name, ".nc")

#Saving raster to disk
writeRaster(grid_raster, file.path("../Spatial_Data/Masks", filename), format = "CDF", overwrite = T,
            varname = "EEZ_regions", zname = "Country_EEZ")
```

# `Python`-based code
We will now start `Python` and save the correct names for the FAO regions in the `netcdf` file we created in `R`.

```{r}
#Activating conda
use_condaenv("CMIP6_data")
```

## Loading libraries

```{python}
import xarray as xr
import numpy as np
import matplotlib.pyplot as plt
import pandas as pd
import os
```

## Loading raster using `xarray`
```{python}
#Getting filename
fn = f'EEZ-world_{r.res}mask{r.mod_name}.nc'

#Loading multilayer raster as dataset
mask = xr.open_dataset(os.path.join('../Spatial_Data/Masks', fn))
#Checking saved file in R
mask
```
Here we can see that the FAO region names are not saved correctly. They are numbered based on its location on the shapefile. We can update this using the data frame with the unique FAO regions we previously created.

```{python}
#We can simply load the data frame from the R environment.
eez_keys = r.eez_names_codes

#Checking list
eez_keys
```
We can now update the names on the `netcdf` file using the ID codes for each EEZ.

```{python}
#We will use the values in the data frame we loaded above
mask['Country_EEZ'] = [int(i) for i in eez_keys.MRGID]
```

We will load the full keys into memory and we will select all EEZ areas for Australia to ensure masks are correct.

```{python}
#Create filename 
#Loading keys
eez_full_keys = pd.read_csv('../Spatial_Data/World_EEZ_v11_20191118/eez_key.csv')

#Selecting Australian subset
aus_eez = mask.sel(Country_EEZ = eez_full_keys['MRGID'][eez_full_keys['SOVEREIGN1'] == 'Australia'].tolist()).EEZ_regions
```


## Plotting results
We will plot all regions below to ensure we got them all correctly.
```{python}
#We will loop through the first five to check results
for reg in aus_eez:
  #Plotting results
  fig = plt.figure()
  ax = fig.add_subplot(111)
  reg.plot(ax = ax, levels = [1, 2])
  plt.title(reg.Country_EEZ.values.tolist())
  plt.show()

```

When comparing to the shapefile plots in the `R` section of this notebook, we can see that the regions are named correctly. This means that we can save our results now.

```{python, eval = F}
#Creating filename
fn = f'EEZ-world-corrected_{r.res}mask{r.mod_name}.nc'

mask.to_netcdf(os.path.join('../Spatial_Data/Masks', fn))
```

## Creating a 2D mask in `csv` format
Depending on the format of your data, it may be better to have a 2D mask in `csv` format. We can create one using the raster we created in the previous step. We will switch to `R` to complete this step.

```{r}
#First we load our raster mask
filename <- paste0("EEZ-world-corrected_", res, "mask", mod_name, ".nc")
# We will load it as a stack
ras <- stack(file.path("../Spatial_Data/Masks", filename), 
             #This is the same variable name we used to create our original raster
             varname = "EEZ_regions")

#We will now convert our stack into points
ras_df <- rasterToPoints(ras) %>% 
  #Transforming form matrix to data frame
  as.data.frame() %>% 
  #Changing dimensions of data frame, so that our column names appear in the `ID_merged column`
  pivot_longer(cols = -c(x, y), names_to = "ID_merged") %>% 
  #We remove any rows containing only NA values
  drop_na(value) %>% 
  #We rename latitude and longitude column
  rename("Lon" = "x", "Lat" = "y") %>% 
  #We remove the `X` from the areas IDs
  mutate(ID_merged = as.integer(str_remove(ID_merged, 'X'))) %>% 
  #Remove the `value` column as it is now redundant
  select(-value)

#We can update our filename variable 
filename <- paste0("EEZ-world_", res, "mask", mod_name, ".csv")
#Saving the end file in your local machine
write_csv(ras_df, file = file.path("../Spatial_Data/Masks", filename))
```
