---
title: "Creating raster mask from FAO regions shapefile"
author: "Denisse Fierro Arcos"
date: "2022-11-07"
output: 
  github_document:
    toc: true
    html_preview: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This notebook will go through the process of creating a raster mask of the FAO major fishing areas available through its [WFS service](https://www.fao.org/fishery/geoserver/fifao/ows?service=WFS&request=GetFeature&version=1.0.0&typeName=fifao:FAO_AREAS_CWP). Rasters are created in `R` and the final details are done with `xarray` library for `Python`.

## Loading R libraries

```{r libraries, results = "hide", warnings = F, message = F}
library(sf)
library(raster)
library(tidyverse)
library(reticulate)
```

## Loading shapefiles

A copy of the FAO Major Fishing Areas was saved locally from the [FAO's WFS service](https://www.fao.org/fishery/geoserver/fifao/ows?service=WFS&request=GetFeature&version=1.0.0&typeName=fifao:FAO_AREAS_CWP).

```{r shapefiles_dirs}
#Load shapefile with FAO regions
fao_reg <- read_sf("../Data/FAO_shapefiles/FAO_MajorAreas.shp") %>% 
  #Subset of columns
  select(-c(F_LEVEL, F_STATUS, SUBOCEAN:F_SUBUNIT, NAME_FR:SURFACE)) %>% 
  #Turning character columns into factors
  mutate_if(is.character, as.factor)

#We can check the results of the first two rows
head(fao_reg, 2)
```

## Plotting FAO regions shapefile
```{r plot_shapefile}
#Loading land shapefile to include in plot
land <- rnaturalearth::ne_countries(type = "countries", returnclass = "sf")
#Plotting FAO shapefile
fao_reg %>% 
  ggplot()+
  geom_sf(aes(fill = NAME_EN))+
  geom_sf(data = land, inherit.aes = F, color = "gray")+
  theme_bw()
```

## Extracting names and codes for FAO regions
We will use this information to save correct names for regions in the raster

```{r fao_keys}
#Create a data frame of unique FAO regions
fao_names_codes <- fao_reg %>% 
  #Tranforming into data frame
  st_drop_geometry() %>% 
  #Extracting unique regions
  distinct(F_CODE, NAME_EN) %>% 
  #Ordering by region code
  arrange(F_CODE) %>% 
  #Fixing up names to use as raster layer name
  mutate_all(as.character) %>% 
  #Removing any spaces and commas from names and replacing with an underscore "_"
  mutate(NAME_EN = str_remove(string = NAME_EN, pattern = ","), 
         NAME_EN = str_replace_all(NAME_EN, " ", "_"))

#We can check some of the results
head(fao_names_codes, 2)
```


## Creating a multilayer raster mask based on merged shapefile
We will now create multilayer mask, which match the resolution of the model forcings (1 deg). First, we will load the sample rasters.

### Loading input rasters
```{r load_rasters_input}
#Loading 1 deg raster to be used as target for rasterising FAO regions
deg1 <- raster("../Data/InputRasters/gfdl-mom6-cobalt2_obsclim_deptho_onedeg_global_fixed.nc")

#Plotting raster
plot(deg1)
```

### Defining function to create rasters from shapefiles
We will define our own function that will use the shapefiles above to create rasters.

```{r function_shp_raster}
#Defining function which needs a shapefile and a raster as input
shp_to_raster <- function(shp, nc_raster){
  #The final raster will have ones where within the shapefile boundaries
  rasterize(shp, nc_raster, field = 1)
}

```

### Applying function to list containing all shapefiles

```{r mapping_function}
#Split shapefile into regions prior to transforming into raster
fao_reg_list <- fao_reg %>% 
  group_by(F_CODE) %>% 
  group_split()

#Applying function to raster list
deg1_raster <- map(fao_reg_list, shp_to_raster, deg1) %>% 
  #Stacking rasters to create a single multilayer raster
  stack()

#Checking results of stacked raster (first six regions)...
plot(deg1_raster[[1:6]])
```
We will plot the shapefile to compare results.

```{r}
#...against shapefile, to make sure they match
fao_reg %>% 
  #Creating a new column with the code and name to identify them easily in graph
  unite("reg_code", F_CODE, NAME_EN, remove = F) %>% 
  ggplot()+
  geom_sf(aes(fill = reg_code))+
  facet_wrap(~F_CODE)
```
The results match! We can now save the results locally and move onto `Python` for the final touches.

```{r}
#Saving raster to disk
writeRaster(deg1_raster, "../Data/Masks/FAO-regions_1degmask.nc", format = "CDF", overwrite = T,
            varname = "FAO_regions", zname = "FAO_reg_name")

#Saving FAO region keys
fao_names_codes %>% 
  write_csv("../Data/Masks/FAO-regions_1degmask_keys.csv")
```

# `Python`-based code
We will now start `Python` and save the correct names for the FAO regions in the `netcdf` file we created in `R`.

```{r}
#Activating conda
use_condaenv(Sys.getenv("RETICULATE_PYTHON"))
```
## Loading libraries

```{python}
import xarray as xr
import numpy as np
import matplotlib.pyplot as plt
import pandas as pd
```
## Loading raster using `xarray`
```{python}
#Loading multilayer raster as dataset
mask = xr.open_dataset("../Data/Masks/FAO-regions_1degmask.nc")
#Checking saved file in R
mask
```
Here we can see that the FAO region names are not saved correctly. They are numbered based on its location on the shapefile. We can update this using the data frame with the unique FAO regions we previously created.

```{python}
#We can simply load the data frame from the R environment.
FAO_keys = r.fao_names_codes
#We could also load it from our disk using the line below.
#FAO_keys = pd.read_csv("../Data/Masks/FAO-regions_1degmask_keys.csv")

#Checking list
FAO_keys
```
We can now update the names on the `netcdf` file.

```{python}
#We will use the values in the data frame we loaded above
mask['FAO_reg_name'] = FAO_keys.NAME_EN.tolist()

#Checking results
mask
```
## Plotting results
We will plot all regions below to ensure we got them all correctly.
```{python}
#We will loop through each layer
for reg in mask.FAO_regions:
  #Plotting results
  fig = plt.figure()
  ax = fig.add_subplot(111)
  reg.plot(ax = ax, levels = [1, 2])
  plt.title(reg.FAO_reg_name.values.tolist())
  plt.show()

```
When comparing to the shapefile plots in the `R` section of this notebook, we can see that the regions are named correctly. This means that we can save our results now.

```{python}
mask.to_netcdf("../Data/Masks/FAO-regions_corrected_1degmask.nc")
```

