---
title: "Creating your own mask from a shapefile"
author: "Denisse Fierro Arcos"
date: "2022-08-30"
output: 
  github_document:
    toc: true
    html_preview: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This notebook will guide you through the steps of how to create a mask using LME shapefiles used in the FishMIP project.

## Loading R libraries

```{r libraries, results = "hide", warnings = F, message = F}
library(sf)
library(raster)
library(tidyverse)
library(stars)
library(geobgu)
```

## Loading shapefiles

LME shapefiles included in the FishMIP project are included under the `Data` folder. First, we will get the file paths for these shapefiles.

### Getting list of LME names from directory paths

```{r shapefiles_dirs}
#Getting the names of directories containing the shapefiles for different regions of interest
names_regions <- list.dirs(path = "../Data/Shapefiles_Regions/", recursive = F, 
                           full.names = F)

#Extracting the names from the directories
names_regions <- str_split_fixed(names_regions, "_", n = 2)[,1]
names_regions
```

### Getting file paths for shapefiles

```{r shapefiles_paths}
#Getting file paths
regions_paths <- list.files(path = "../Data/Shapefiles_Regions/", pattern = ".shp$", 
                            recursive = T, full.names = T)

#Ignore any shapefiles included in the "SupportInfo" subfolder
regions_paths <- regions_paths[!str_detect(regions_paths, "SupportInfo")]
regions_paths
```

### Loading regions

We must first ensure that all shapefiles are in the same coordinate reference system (CRS) prior to creating a single shapefile. We will load each shapefile into a list and change the CRS if necessary. We are also making sure that any shapefiles that cross the international dateline are shown correctly when WGS84 CRS (EPSG:4326) is assigned.

```{r shapefiles_load}
#Switching off spherical geometry prior to accessing and cropping world shapefile
sf_use_s2(F)

#Create empty list to store shapefiles
AOI_list <- list()

for(i in seq_along(regions_paths)){
  #Load original shapefile
  reg_raw <- read_sf(regions_paths[i]) %>%
    mutate(region = names_regions[i])
  #Check shapefile CRS. If none assigned and geometry is between -180 and +180, then assign WGS84
  if(st_crs(reg_raw) == "" | is.na(st_crs(reg_raw)) | is.null(st_crs(reg_raw))){
    if(sum(reg_raw$geometry[[1]][[1]] > 180) == F & sum(reg_raw$geometry[[1]][[1]] < -180) == F){
      st_crs(reg_raw) <- 4326}
    #If the shapefile has a CRS different to 4326, then transform it to WGS84
    }else if(st_crs(reg_raw) != st_crs(4326)){
      reg_raw <- st_transform(reg_raw, 4326)
    }
  
  #Remove internal region boundaries
  reg_raw <- reg_raw %>% 
    #Make sure it wraps around the international dateline
    st_wrap_dateline(options = c("WRAPDATELINE=YES", "DATELINEOFFSET=180")) %>% 
    group_by(region) %>% 
    summarise() %>% 
    #Ensure there are only two dimensions
    st_zm()
  
  #Make sure final area is polygon only
  if(unique(st_geometry_type(reg_raw)) != "POLYGON"){
    reg_raw <- st_cast(reg_raw, "POLYGON", do_split = F)
  }
  
  #Saving shapefiles in empty list
  AOI_list[[names_regions[i]]] <- reg_raw
}
```
Now that all shapefiles are in the same CRS, we can create a single shapefile for all LME regions.

```{r shapefiles_merge}
RMEs <- AOI_list %>% 
  bind_rows()
```

## Plotting merged shapefile
```{r plot_shapefile}
#We will load a world map for reference
land <- rnaturalearth::ne_countries(type = "countries", returnclass = "sf")

#Plotting LMEs over world map
RMEs %>% 
  ggplot()+
  geom_sf(aes(fill = region))+
  geom_sf(data = land, inherit.aes = F, color = "gray")+
  theme_bw()
```

### Saving LME polygons as a single shapefile

```{r shapefile_saving}
st_write(RMEs, "../Data/Masks/FishMIP_RMEs_all.shp", append = F)
```


## Creating a multilayer raster mask based on merged shapefile
We will now create multilayer masks at two resolutions: one degree and 0.25 degrees, which match the resolution of the model forcings. We will first load sample rasters at these two resolutions.

### Loading input rasters
```{r load_rasters_input}
deg025 <- raster("../Data/InputRasters/gfdl-mom6-cobalt2_obsclim_deptho_15arcmin_global_fixed.nc")
deg1 <- raster("../Data/InputRasters/gfdl-mom6-cobalt2_obsclim_deptho_onedeg_global_fixed.nc")

#Plotting raster
plot(deg1)
```

### Defining function to create rasters from shapefiles
We will define our own function that will use the shapefiles above to create rasters.

```{r function_shp_raster}
#Defining function which needs a shapefile and a raster as input
shp_to_raster <- function(shp, nc_raster){
  #If needed, sf shapefiles can be transformed into Spatial objects
  # shp <- as(shp, "Spatial")
  #The final raster will have ones where within the shapefile boundaries
  rasterize(shp, nc_raster, field = 1)
}

```

### Applying function to list containing all shapefiles

```{r mapping_function}
#Applying function defined above to all shapefiles within list
deg1_raster <- map(AOI_list, shp_to_raster, deg1)
deg025_raster <- map(AOI_list, shp_to_raster, deg025)

#Stacking rasters to create a single multilayer raster for each resolution
deg1_raster <- stack(deg1_raster)
deg025_raster <- stack(deg025_raster)

```

### Saving rasters to disk

```{r}
writeRaster(deg025_raster, "../Data/Masks/fishMIP_regional_025mask_ISIMIP3a.nc", format = "CDF", overwrite = T,
            xname = "Longitude", yname = "Latitude", varname = "RegionMask", varunit = "binary",
            longname = "Region Mask -- True or False", zname = "Region")

writeRaster(deg1_raster, "../Data/Masks/fishMIP_regional_1mask_ISIMIP3a.nc", format = "CDF", overwrite = T,
            xname = "Longitude", yname = "Latitude", varname = "RegionMask", varunit = "binary",
            longname = "Region Mask -- True or False", zname = "Region")
```
## Using Python to update name of regions in netcdf file

```{r py_ret}
library(reticulate)
python_path <- as.character(Sys.which("python"))
use_python(python_path)
```


```{python}
#Loading xarray library to open netcdf file
import xarray as xr

#Open stacked rasters
deg1 = xr.open_dataset("../Data/Masks/fishMIP_regional_1mask_ISIMIP3a.nc")
deg025 = xr.open_dataset("../Data/Masks/fishMIP_regional_025mask_ISIMIP3a.nc")

RME_names = r.names_regions

#Create an empty dictionary
RME_mask1deg = []
RME_mask025deg = []

#Loop through each dictionary entry
for da1deg, da025deg, rme in zip(deg1.RegionMask, deg025.RegionMask, RME_names):
  da1deg = da1deg.drop_vars("Region")
  da025deg = da025deg.drop_vars("Region")
  #Adding dimension with LME name
  RME_mask1deg.append(da1deg.expand_dims(RME_name = [rme]))
  RME_mask025deg.append(da025deg.expand_dims(RME_name = [rme]))

#Creating multidimensional dataset
RME_mask1deg = xr.concat(RME_mask1deg, dim = 'RME_name')
RME_mask025deg = xr.concat(RME_mask025deg, dim = 'RME_name')

#Check results
RME_mask1deg; RME_mask025deg

RME_mask1deg.to_netcdf("../Data/Masks/fishMIP_regional_1degmask_ISIMIP3a.nc", mode = 'w')
RME_mask025deg.to_netcdf("../Data/Masks/fishMIP_regional_025degmask_ISIMIP3a.nc", mode = 'w')
```
### Checking results in R

```{r plot_stack}
deg1 <- stack("../Data/Masks/fishMIP_regional_1degmask_ISIMIP3a.nc", varname = "RegionMask")
deg025 <- stack("../Data/Masks/fishMIP_regional_025degmask_ISIMIP3a.nc", varname = "RegionMask")

plot(deg1)
```

## Extracting data with multipolygon feature
Finally, we will use the newly created shapefile to extract data for our regions of interest.

```{r extracting_data_raster}
#Getting file paths of raster with global total catch data
data_file <- list.files(path = "../Data/", pattern = ".*global.*nc$", full.names = T)

#Loading sample raster from disk
tc_raster <- read_stars(data_file)%>% 
  st_set_crs(st_crs(RMEs))

#Change name of variable for easier access to data
names(tc_raster) <- "tc"

max_val <- max(tc_raster$tc)

#Masking land values
tc_raster <- st_apply(tc_raster, "time", function(x) na_if(x, as.numeric(max_val)))
```

Plotting one time step as an example.

```{r raster_slice}
tc1 <- tc_raster %>%
  slice(index = 1, along = "time")
plot(tc1)
```


## Extract data from raster

### Option A with `st_crop`
```{r raster_crop}
tc1_crop <- st_crop(tc1, RMEs)
plot(tc1_crop)
```

### Option B with simple mask
```{r raster_mask}
tc1_crop2 <- tc1[RMEs]
plot(tc1_crop2)
```

## Extracting time series
Get the dates for time steps in raster
```{r ts_dates}
time_steps <- st_get_dimension_values(tc_raster, "time")
head(time_steps)
```
## Plotting time series
Extracting mean per timestep. Showing timeseries for one region only.

```{r ts_plot}
#Extracting data for all timesteps
rme_extract <- RMEs %>% 
  mutate(mean = raster_extract(tc_raster, RMEs, fun = mean, na.rm = T))

#Transforming shapefile into dataframe
rme_extract <- rme_extract %>% 
  st_drop_geometry()

#Shaping dataframe better before plotting
ts_lme <- as.data.frame(rme_extract$mean)
names(ts_lme) <- time_steps
ts_lme <- ts_lme %>% 
  mutate(region = rme_extract$region, .before = 1) %>% 
  pivot_longer(!region, names_to = "date", values_to = "mean_tc") %>% 
  mutate(date = lubridate::ymd(date),
         region = factor(region))

#Plotting results
ts_lme %>%
  ggplot(aes(date, mean_tc))+
  geom_line(aes(colour = region))
```


